Combinatorial Mixture Models for Single Cell Assays with Application to Vaccine Studies

### 2012-07-19 MIMOSA.tex
Updated multivariate simulations
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions. The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (Figure~\ref{fig:simulations_trunc}). Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test and performed about as well as the constrained MIMOSA model fit to one-sided data.

@@

The MIMOSA model presented here uses a mixture model framework of Beta--Binomial or Multinomial--Dirichlet distributions to model counts in experimental units (i.e. individuals) across multiple conditions (i.e. vaccine responders and non--responders). Information is shared across non--responder individuals and across responders individuals through exchangeable Beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when model assumptions are violated (Figures~\ref{fig:simulations} and~\ref{fig:simulations_trunc}). The MIMOSA model based on the Beta--Binomial distribution allows us to constrain the alternative hypothesis to the case $p_s > p_u$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample.

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=0.3\columnwidth]{Figures/Sim_Trunc_AUC.pdf}};
    \node[anchor=south west, inner sep=0] at (4.5,0) {\includegraphics[width=0.3\columnwidth]{Figures/Sim_Trunc_ROC.pdf}};
   \node[anchor=south west, inner sep=0] at (9,0) {\includegraphics[width=0.3\columnwidth]{Figures/Sim_Trunc_FDR.pdf}};
    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
    \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
    \node at (9,3.4) [font=\tiny\sffamily] {C} ;
 \end{tikzpicture}
   \caption{Unconstrained MIMOSA model fit to data where model assumptions are violated. Two--sided data were simulated from a model where the proportions were drawn from a truncated normal distribution over $[0,1]$, rather than a Beta distribution.}
   \label{fig:simulations_trunc}
\end{figure}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.3\columnwidth]{Figures/Multivariate_Sims_AUCBOX.pdf}};
    \node[anchor=south west, inner sep=0] at (4.5,0){\includegraphics[width=0.3\columnwidth]{Figures/Multivariate_Sims_FDR.pdf}};
   \node[anchor=south west, inner sep=0] at (9,0) {\includegraphics[width=0.3\columnwidth]{Figures/Multivariate_Sims_ROC.pdf}};
    \node at (0,2.1) [font=\tiny\sffamily] {A} ;
    \node at (4.5,2.1) [font=\tiny\sffamily] {B} ;
    \node at (9,2.1) [font=\tiny\sffamily] {C} ;
\end{tikzpicture}
   \caption{Multivariate simulations from a two--sided model. Ten, five-dimensional data sets were simulated from a two--sided model with an effect sizes of $2.5\times 10^{-3}$ and $-2.5\times 10^{-3}$ in two of the five dimensions. Multivariate MIMOSA was compared against Fisher's exact test, the likelihood ratio test, and Fisher's combined p--value, combining Fisher's exact test run marginally on each of the five dimensions. A) Boxplots of AUCs (area under the curve) for the ten simulations. B) Observed and nominal false discovery rate for each method. C)ROC curves for the competing methods.}
   \label{fig:mvsimulations}
\end{figure}
\clearpage
%
%\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
%   \centering
%   \includegraphics[width=4in]{Figures/simulations_violatedmodel} 
%   \caption{Sensitivity to departures from model assumptions. We generated data from a variant of the constrained model ($p_s>p_u$) where the proportions were simulated from truncated normal distributions on $(0,1)$, rather than from beta distributions. The mean and variance of the normal distributions was given by $\mu=\alpha/(\alpha+beta), \sigma^2=(\alpha\beta)/((\alpha+beta)^2(\alpha+\beta+1))$, where $\alpha,\beta$ are the Beta--prior hyper parameters for the MIMOSA model estimated from real data ($s, u$ subscripts omitted for brevity). The AUC for the unconstrained MIMOSA model and Fisher's exact test as a function of event count are shown in the first panel. The beta distributions for the estimated and true hyper parameters  are shown in the second panel. The ROC curves for one data set and the observed and true false discovery rates for the model and Fisher's exact test are shown in the third and fourth panels.}
%   \label{fig:simulations_trunc}
%\end{figure}

@@

>>>>

To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions. The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (Supplementary Figure~\ref{fig:simulations_trunc}). Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test and performed about as well as the constrained MIMOSA model fit to one-sided data.

@@

The MIMOSA model presented here uses a mixture model framework of Beta--Binomial or Multinomial--Dirichlet distributions to model counts in experimental units (i.e. individuals) across multiple conditions (i.e. vaccine responders and non--responders). Information is shared across non--responder individuals and across responders individuals through exchangeable Beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when model assumptions are violated (Figures~\ref{fig:simulations} and Supplementary Figure~\ref{fig:simulations_trunc}). The MIMOSA model based on the Beta--Binomial distribution allows us to constrain the alternative hypothesis to the case $p_s > p_u$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample.

@@

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0){\includegraphics[width=0.4\columnwidth]{Figures/Multivariate_Sims_ROC.pdf}};
   \node[anchor=south west, inner sep=0] at (6,0) {\includegraphics[width=0.4\columnwidth]{Figures/Multivariate_Sims_FDR.pdf}};
    \node at (0,3.5) [font=\tiny\sffamily] {A} ;
    \node at (6,3.5) [font=\tiny\sffamily] {B} ;
\end{tikzpicture}
   \caption{Multivariate simulations from a two--sided model. Ten, eight-dimensional data sets were simulated from a two--sided model with an effect sizes of $2.5\times 10^{-3}$ and $-2.5\times 10^{-3}$ in two of the eight dimensions (N=1,500). Multivariate MIMOSA was compared against Fisher's exact test, the likelihood ratio test, and Fisher's combined p--value (combining Fisher's exact test run marginally on each of the five dimensions). A) Average ROC curves for the competing methods over 10 simulations. B) Average observed and nominal false discovery rate for each method over 10 simulations.}
   \label{fig:mvsimulations}
\end{figure}
\clearpage
%
%\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
%   \centering
%   \includegraphics[width=4in]{Figures/simulations_violatedmodel} 
%   \caption{Sensitivity to departures from model assumptions. We generated data from a variant of the constrained model ($p_s>p_u$) where the proportions were simulated from truncated normal distributions on $(0,1)$, rather than from beta distributions. The mean and variance of the normal distributions was given by $\mu=\alpha/(\alpha+beta), \sigma^2=(\alpha\beta)/((\alpha+beta)^2(\alpha+\beta+1))$, where $\alpha,\beta$ are the Beta--prior hyper parameters for the MIMOSA model estimated from real data ($s, u$ subscripts omitted for brevity). The AUC for the unconstrained MIMOSA model and Fisher's exact test as a function of event count are shown in the first panel. The beta distributions for the estimated and true hyper parameters  are shown in the second panel. The ROC curves for one data set and the observed and true false discovery rates for the model and Fisher's exact test are shown in the third and fourth panels.}
%   \label{fig:simulations_trunc}
%\end{figure}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_5000.pdf}};
    \node[anchor=south west, inner sep=0] at (6,0) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_5000.pdf}};
    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
    \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
 \end{tikzpicture}
   \caption{Unconstrained MIMOSA model fit to data where model assumptions are violated. Two--sided data were simulated from a model where the proportions were drawn from a truncated normal distribution over $[0,1]$, rather than a Beta distribution. A) Average ROC curves from 10 simulations and B) average observed vs nominal FDR from 10 simulations at N=5,000.}
   \label{fig:simulations_trunc}
\end{figure}
